\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-04-16 analysis/paper/paper.tex
Write up and plots fixed. plot descriptions fixed
All tasks complete. I will need to do a last grammar/writing check

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: elastic-net penalization with
\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
\citeyearpar{clements_liu}, while \pkg{survival}
\citeyearpar{survival-package} has an implementation for ridge.
\pkg{casebase} provides penalized estimation in the context of
parametric hazards. To our knowledge, \pkg{casebase} and \pkg{rsptm2}
are the only packages to offer this functionality.
\textcolor{red}{I can dive deeper to determine if this is a fair statement}

@@

\begin{CodeInput}
R> data(ERSPC)
R> ERSPC$ScrArm <- factor(ERSPC$ScrArm,
R>   levels = c(0, 1),
R>   labels = c("Control group", "Screening group")
R> )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> pt_object_strat <- casebase::popTime(ERSPC,
R>   event = "DeadOfPrCa",
R>   exposure = "ScrArm"
R> )
R> plot(pt_object_strat)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
R>   data = ERSPC
R> )
R> summary(cox_model)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> new_data <- data.frame(
R>   ScrArm = c("Control group", "Screening group"),
R>   ignore = 99
R> )
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>   xlab = "Years since Randomization",
R>   ylab = "Cumulative Incidence (%)",
R>   fun = "event",
R>   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
R>   main = sprintf(
R>     "Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>     exp(coef(cox_model)),
R>     exp(confint(cox_model))[1],
R>     exp(confint(cox_model))[2]
R>   ),
R>   ylim = c(0, 2), yscale = 100
R> )
R> legend("topleft",
R>   legend = c("Control group", "Screening group"),
R>   col = c("red", "blue"),
R>   lty = c(1, 1),
R>   bg = "gray90"
R> )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm,
R>   data = ERSPC,
R>   ratio = 100
R> )
R> 
R> summary(casebase_exponential)
R> exp(coef(casebase_exponential)[2])
R> exp(confint(casebase_exponential)[2, ])
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_exp <- casebase::absoluteRisk(
R>   object = casebase_exponential,
R>   time = seq(0, 15, 0.1),
R>   newdata = new_data
R> )
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>   xlab = "Years since Randomization",
R>   ylab = "Cumulative Incidence (%)",
R>   fun = "event",
R>   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
R>   main = sprintf(
R>     "Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>     exp(coef(cox_model)),
R>     exp(confint(cox_model))[1],
R>     exp(confint(cox_model))[2]
R>   ),
R>   ylim = c(0, 2), yscale = 100
R> )
R> lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
R> lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)
R> 
R> 
R> legend("topleft",
R>   legend = c(
R>     "Control group (Cox)", "Control group (Casebase)",
R>     "Screening group (Cox)", "Screening group (Casebase)"
R>   ),
R>   col = c("red", "red", "blue", "blue"),
R>   lty = c(1, 2, 1, 2),
R>   bg = "gray90"
R> )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm,
R>   data = ERSPC,
R>   ratio = 100
R> )
R> 
R> summary(casebase_time)
R> exp(coef(casebase_time))
R> exp(confint(casebase_time))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_time <- casebase::absoluteRisk(
R>   object = casebase_time,
R>   time = seq(0, 15, 0.1),
R>   newdata = new_data
R> )
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>   xlab = "Years since Randomization",
R>   ylab = "Cumulative Incidence (%)",
R>   fun = "event",
R>   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
R>   main = sprintf(
R>     "Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>     exp(coef(cox_model)),
R>     exp(confint(cox_model))[1],
R>     exp(confint(cox_model))[2]
R>   ),
R>   ylim = c(0, 2), yscale = 100
R> )
R> lines(smooth_risk_time[, 1], smooth_risk_time[, 2], col = "red", lty = 2)
R> lines(smooth_risk_time[, 1], smooth_risk_time[, 3], col = "blue", lty = 2)
R> 
R> legend("topleft",
R>   legend = c(
R>     "Control group (Cox)", "Control group (Casebase)",
R>     "Screening group (Cox)", "Screening group (Casebase)"
R>   ),
R>   col = c("red", "red", "blue", "blue"),
R>   lty = c(1, 2, 1, 2),
R>   bg = "gray90"
R> )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm,
R>   data = ERSPC,
R>   ratio = 100
R> )
R> 
R> summary(casebase_splines)
R> exp(coef(casebase_splines))
R> exp(confint(casebase_splines))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_splines <- absoluteRisk(
R>   object = casebase_splines,
R>   time = seq(0, 15, 0.1),
R>   newdata = new_data
R> )
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>   xlab = "Years since Randomization",
R>   ylab = "Cumulative Incidence (%)",
R>   fun = "event",
R>   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
R>   main = sprintf(
R>     "Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>     exp(coef(cox_model)),
R>     exp(confint(cox_model))[1],
R>     exp(confint(cox_model))[2]
R>   ),
R>   ylim = c(0, 2), yscale = 100
R> )
R> lines(smooth_risk_splines[, 1], smooth_risk_splines[, 2], col = "red", lty = 2)
R> lines(smooth_risk_splines[, 1], smooth_risk_splines[, 3], col = "blue", lty = 2)
R> 
R> legend("topleft",
R>   legend = c(
R>     "Control group (Cox)", "Control group (Casebase)",
R>     "Screening group (Cox)", "Screening group (Casebase)"
R>   ),
R>   col = c("red", "red", "blue", "blue"),
R>   lty = c(1, 2, 1, 2),
R>   bg = "gray90"
R> )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> c(
R>   "Exp." = AIC(casebase_exponential),
R>   "Weib." = AIC(casebase_time),
R>   "Splines" = AIC(casebase_splines)
R> )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> model_cb <- fitSmoothHazard(
R>   Status ~ ftime + Sex + D + Phase + Source + Age,
R>   data = bmtcrr,
R>   ratio = 100,
R>   time = "ftime"
R> )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(survival)
R> # Treat competing event as censoring
R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
R>   data = bmtcrr
R> )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Pick 100 equidistant points between 0 and 60 months
R> time_points <- seq(0, 60, length.out = 50)
R> 
R> # Data.frame containing risk profile
R> newdata <- data.frame(
R>   "Sex" = factor(c("F", "F"),
R>     levels = levels(bmtcrr[, "Sex"])
R>   ),
R>   "D" = c("ALL", "AML"), # Both diseases
R>   "Phase" = factor(c("Relapse", "Relapse"),
R>     levels = levels(bmtcrr[, "Phase"])
R>   ),
R>   "Age" = c(35, 35),
R>   "Source" = factor(c("PB", "PB"),
R>     levels = levels(bmtcrr[, "Source"])
R>   )
R> )
R> 
R> # Estimate absolute risk curve
R> risk_cb <- absoluteRisk(
R>   object = model_cb, time = time_points,
R>   method = "numerical", newdata = newdata
R> )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(timereg)
R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
R>   const(Phase) + const(Source) + const(Age),
R> data = bmtcrr, cause = 1, model = "fg"
R> )
R> 
R> # Estimate absolute risk curve
R> risk_fg <- predict(model_fg, newdata, times = time_points)
\end{CodeInput}
\end{CodeChunk}

@@

In the first two case studies, we described the basic functionalities of
the \pkg{casebase} package: creating population-time plots, fitting
parametric models for hazard functions, and estimating the corresponding
cumulative incidence curves. In the next two case studies, we explore
more advanced features.

For the third case study, we will look at regularized estimation for the
hazard function. This can be performed using a combination of case-base
sampling and regularized logistic regression. To demonstrate this
capability, we examine the Study to Understand Prognoses Preferences
Outcomes and Risks of Treatment (SUPPORT) dataset
\citep{knaus1995support}. The SUPPORT dataset tracks death for
individuals who are considered seriously ill within 5 hospitals in the
United States of America. Imputation was conducted using the \pkg{mice}
package in \proglang{R} with default settings \citep{mice}. Before
imputation, there were a total of 9105 individuals and 47 variables. Of
those variables, a few were removed before imputation. We removed three
response variables: hospital charges, patient ratio of costs to charges
and patient micro-costs. Next, we removed hospital death as it was
directly informative of our event of interest, death. We also removed
functional disability and income as they are ordinal covariates, because
ordinal variables are treated as factors in our current model. Finally,
we removed 8 covariates relating to the results of previous findings: we
removed SUPPORT day 3 physiology score (sps), APACHE III day 3
physiology score (aps), SUPPORT model 2-month survival estimate, SUPPORT
model 6-month survival estimate, Physician's 2-month survival estimate
for pt., Physician's 6-month survival estimate for pt., Patient had Do
Not Resuscitate (DNR) order and Day of DNR order (\textless{}0 if before
study). Of these, sps and aps were added on after imputation, as they
were missing only 1 observation and permitted us to compare findings
between our model and their's.

First we impute manually using recommended normal values for
physiological measures \citep{harrell_2020}. Next, we impute a single
dataset using \pkg{mice} with default settings. After imputation, we
noted that the covariate for surrogate activities of daily living was
not imputed. This is due to colinearity between the other two covariates
for activities of daily living. As such, surrogate activites of daily
living was removed from our model. After these imputations, we have a
total of 9104 observations and 30 variables, two of which are sps and
aps, which will work as a baseline for our comparison to previous
findings. This dataset is available Its worth noting that these scores
were developed for a period of 180 days, which makes our comparison over
5.56 years exploratory. Also, our single imputation introduces bias in
the data. As the goal here is purely demonstrative, this bias will be
consistent with any of the following models. As such, their comparison
is still valid, though any interpretation requires an awareness of said
bias. A description of used variables can be found in Table
\ref{tab:support1} and a breakdown of each categorical variable in Table
\ref{tab:support2}. The original data was obtained from the Department
of Biostatistics at Vanderbilt University \citep{harrell_2020}. The
imputed dataset is available as part of the \pkg{casebase} package.

@@

As before, the first step is to visualize the incidence density. The
population-time plot for the SUPPORT data appears in Figure
\ref{fig:support-poptime}. As we can see, the incidence density is not
constant.

{\centering \includegraphics{../figures/support-poptime-1}

\caption{\label{fig:support-poptime} Population-time plot for the SUPPORT data.}\label{fig:support-poptime}
\end{figure}
\end{CodeChunk}

Before generating an absolute risk curve using regularized logistic
regression, we will explore three options as a baseline comparison,
beyond a Kaplan-meier curve. The original study compared these scores
over a span of 180 days, which may not be comparable over the full 5.56
years. Therefore, We will examine three cox regression scenarios: sps
only, aps only and all covariates excluding sps and aps. Next, we will
compare each modeling using the Brier score \citep{graf1999ass}. The
Brier score is able to demonstrate both discrimination and calibration
in a single model \citep{graf1999ass} and provide a comparable metric
between semi-parametric and parametric methods in the survival context.
For all the following analyses in this case study, we will use 80\% of
the observations to train our models, while 20\% will be used to
generate absolute risk curves, all of which will be averaged to
demonstrate each model's predictive ability. Over 5.56 years,
\ref{fig:baselines} demonstrates that all the cox models have similar,
if not identical average predictions. Figure \ref{fig:brier1} examines
the brier scores as they change over time. As the full model has the
lowest brier score on overall, it will be used in the following
comparison to regularized cox, regularized logistic regression and the
unadjusted Kaplan-Meier curve.

@@

>>>>

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: elastic-net penalization with
\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
\citeyearpar{clements_liu}, while \pkg{survival}
\citeyearpar{survival-package} has an implementation for ridge.
\pkg{casebase} provides penalized estimation in the context of
parametric hazards. To our knowledge, \pkg{casebase} and \pkg{rsptm2}
are the only packages to offer this functionality.

@@

\begin{CodeInput}
R> data(ERSPC)
R> ERSPC$ScrArm <- factor(ERSPC$ScrArm,
R+   levels = c(0, 1),
R+   labels = c("Control group", "Screening group")
R+ )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> pt_object_strat <- casebase::popTime(ERSPC,
R+   event = "DeadOfPrCa",
R+   exposure = "ScrArm"
R+ )
R> plot(pt_object_strat)
\end{CodeInput}

\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-1} \end{center}

@@

\begin{CodeInput}
R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
R+   data = ERSPC
R+ )
R> summary(cox_model)
\end{CodeInput}

\begin{CodeOutput}
#> Call:
#> survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
#>     ScrArm, data = ERSPC)
#> 
#>   n= 159893, number of events= 540 
#> 
#>                         coef exp(coef) se(coef)     z Pr(>|z|)  
#> ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#>                       exp(coef) exp(-coef) lower .95 upper .95
#> ScrArmScreening group     0.801       1.25     0.674     0.952
#> 
#> Concordance= 0.519  (se = 0.011 )
#> Likelihood ratio test= 6.45  on 1 df,   p=0.01
#> Wald test            = 6.37  on 1 df,   p=0.01
#> Score (logrank) test = 6.39  on 1 df,   p=0.01
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> new_data <- data.frame(
R+   ScrArm = c("Control group", "Screening group"),
R+   ignore = 99
R+ )
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+   xlab = "Years since Randomization",
R+   ylab = "Cumulative Incidence (%)",
R+   fun = "event",
R+   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
R+   main = sprintf(
R+     "Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+     exp(coef(cox_model)),
R+     exp(confint(cox_model))[1],
R+     exp(confint(cox_model))[2]
R+   ),
R+   ylim = c(0, 2), yscale = 100
R+ )
R> legend("topleft",
R+   legend = c("Control group", "Screening group"),
R+   col = c("red", "blue"),
R+   lty = c(1, 1),
R+   bg = "gray90"
R+ )
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-cox-cif-1} \end{center}

@@

\begin{CodeInput}
R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm,
R+   data = ERSPC,
R+   ratio = 100
R+ )
R> 
R> summary(casebase_exponential)
\end{CodeInput}

\begin{CodeOutput}
#> 
#> Call:
#> glm(formula = formula, family = binomial, data = sampleData)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.148  -0.148  -0.148  -0.133   3.078  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value            Pr(>|z|)    
#> (Intercept)            -7.7659     0.0557 -139.46 <0.0000000000000002 ***
#> ScrArmScreening group  -0.2147     0.0884   -2.43               0.015 *  
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059  on 54539  degrees of freedom
#> Residual deviance: 6053  on 54538  degrees of freedom
#> AIC: 6057
#> 
#> Number of Fisher Scoring iterations: 7
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_exponential)[2])
\end{CodeInput}

\begin{CodeOutput}
#> ScrArmScreening group 
#>                  0.81
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_exponential)[2, ])
\end{CodeInput}

\begin{CodeOutput}
#>  2.5 % 97.5 % 
#>   0.68   0.96
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_exp <- casebase::absoluteRisk(
R+   object = casebase_exponential,
R+   time = seq(0, 15, 0.1),
R+   newdata = new_data
R+ )
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+   xlab = "Years since Randomization",
R+   ylab = "Cumulative Incidence (%)",
R+   fun = "event",
R+   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
R+   main = sprintf(
R+     "Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+     exp(coef(cox_model)),
R+     exp(confint(cox_model))[1],
R+     exp(confint(cox_model))[2]
R+   ),
R+   ylim = c(0, 2), yscale = 100
R+ )
R> lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
R> lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)
R> 
R> 
R> legend("topleft",
R+   legend = c(
R+     "Control group (Cox)", "Control group (Casebase)",
R+     "Screening group (Cox)", "Screening group (Casebase)"
R+   ),
R+   col = c("red", "red", "blue", "blue"),
R+   lty = c(1, 2, 1, 2),
R+   bg = "gray90"
R+ )
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-casebase-cif-1} \end{center}

@@

\begin{CodeInput}
R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm,
R+   data = ERSPC,
R+   ratio = 100
R+ )
R> 
R> summary(casebase_time)
\end{CodeInput}

\begin{CodeOutput}
#> 
#> Call:
#> glm(formula = formula, family = binomial, data = sampleData)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.401  -0.162  -0.122  -0.094   3.466  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value            Pr(>|z|)    
#> (Intercept)            -9.0314     0.1121  -80.53 <0.0000000000000002 ***
#> Follow.Up.Time          0.2222     0.0145   15.29 <0.0000000000000002 ***
#> ScrArmScreening group  -0.2320     0.0886   -2.62              0.0089 ** 
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059  on 54539  degrees of freedom
#> Residual deviance: 5811  on 54537  degrees of freedom
#> AIC: 5817
#> 
#> Number of Fisher Scoring iterations: 8
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_time))
\end{CodeInput}

\begin{CodeOutput}
#>           (Intercept)        Follow.Up.Time ScrArmScreening group 
#>               0.00012               1.24882               0.79297
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_time))
\end{CodeInput}

\begin{CodeOutput}
#>                          2.5 %  97.5 %
#> (Intercept)           0.000096 0.00015
#> Follow.Up.Time        1.213876 1.28507
#> ScrArmScreening group 0.665769 0.94256
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_time <- casebase::absoluteRisk(
R+   object = casebase_time,
R+   time = seq(0, 15, 0.1),
R+   newdata = new_data
R+ )
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+   xlab = "Years since Randomization",
R+   ylab = "Cumulative Incidence (%)",
R+   fun = "event",
R+   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
R+   main = sprintf(
R+     "Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+     exp(coef(cox_model)),
R+     exp(confint(cox_model))[1],
R+     exp(confint(cox_model))[2]
R+   ),
R+   ylim = c(0, 2), yscale = 100
R+ )
R> lines(smooth_risk_time[, 1], smooth_risk_time[, 2], col = "red", lty = 2)
R> lines(smooth_risk_time[, 1], smooth_risk_time[, 3], col = "blue", lty = 2)
R> 
R> legend("topleft",
R+   legend = c(
R+     "Control group (Cox)", "Control group (Casebase)",
R+     "Screening group (Cox)", "Screening group (Casebase)"
R+   ),
R+   col = c("red", "red", "blue", "blue"),
R+   lty = c(1, 2, 1, 2),
R+   bg = "gray90"
R+ )
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-casebase-weibull-cif-1} \end{center}

@@

\begin{CodeInput}
R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm,
R+   data = ERSPC,
R+   ratio = 100
R+ )
R> 
R> summary(casebase_splines)
\end{CodeInput}

\begin{CodeOutput}
#> 
#> Call:
#> glm(formula = formula, family = binomial, data = sampleData)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.405  -0.173  -0.136  -0.085   3.796  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value             Pr(>|z|)    
#> (Intercept)           -10.2643     0.3159  -32.49 < 0.0000000000000002 ***
#> bs(Follow.Up.Time)1     4.2983     0.8005    5.37        0.00000007893 ***
#> bs(Follow.Up.Time)2     2.0945     0.4824    4.34        0.00001412028 ***
#> bs(Follow.Up.Time)3     4.5544     0.7067    6.44        0.00000000012 ***
#> ScrArmScreening group  -0.2169     0.0886   -2.45                0.014 *  
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059.0  on 54539  degrees of freedom
#> Residual deviance: 5790.5  on 54535  degrees of freedom
#> AIC: 5800
#> 
#> Number of Fisher Scoring iterations: 8
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_splines))
\end{CodeInput}

\begin{CodeOutput}
#>           (Intercept)   bs(Follow.Up.Time)1   bs(Follow.Up.Time)2 
#>              0.000035             73.578176              8.121029 
#>   bs(Follow.Up.Time)3 ScrArmScreening group 
#>             95.053534              0.805017
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_splines))
\end{CodeInput}

\begin{CodeOutput}
#>                           2.5 %     97.5 %
#> (Intercept)            0.000018   0.000063
#> bs(Follow.Up.Time)1   15.835008 367.061126
#> bs(Follow.Up.Time)2    3.252368  21.652009
#> bs(Follow.Up.Time)3   22.537699 364.141700
#> ScrArmScreening group  0.675920   0.956823
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_splines <- absoluteRisk(
R+   object = casebase_splines,
R+   time = seq(0, 15, 0.1),
R+   newdata = new_data
R+ )
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+   xlab = "Years since Randomization",
R+   ylab = "Cumulative Incidence (%)",
R+   fun = "event",
R+   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
R+   main = sprintf(
R+     "Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+     exp(coef(cox_model)),
R+     exp(confint(cox_model))[1],
R+     exp(confint(cox_model))[2]
R+   ),
R+   ylim = c(0, 2), yscale = 100
R+ )
R> lines(smooth_risk_splines[, 1], smooth_risk_splines[, 2], col = "red", lty = 2)
R> lines(smooth_risk_splines[, 1], smooth_risk_splines[, 3], col = "blue", lty = 2)
R> 
R> legend("topleft",
R+   legend = c(
R+     "Control group (Cox)", "Control group (Casebase)",
R+     "Screening group (Cox)", "Screening group (Casebase)"
R+   ),
R+   col = c("red", "red", "blue", "blue"),
R+   lty = c(1, 2, 1, 2),
R+   bg = "gray90"
R+ )
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-casebase-splines-cif-1} \end{center}

@@

\begin{CodeInput}
R> c(
R+   "Exp." = AIC(casebase_exponential),
R+   "Weib." = AIC(casebase_time),
R+   "Splines" = AIC(casebase_splines)
R+ )
\end{CodeInput}

\begin{CodeOutput}
#>    Exp.   Weib. Splines 
#>    6057    5817    5800
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> model_cb <- fitSmoothHazard(
R+   Status ~ ftime + Sex + D + Phase + Source + Age,
R+   data = bmtcrr,
R+   ratio = 100,
R+   time = "ftime"
R+ )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(survival)
R> # Treat competing event as censoring
R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
R+   data = bmtcrr
R+ )
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{table}

\caption{\label{tab:bmtcrr-cis}Estimates and confidence intervals for the hazard ratios for each coefficient. Both estimates from case-base sampling and Cox regression are presented.}
\centering
\begin{tabular}[t]{lrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Covariates & HR & 95\% CI & HR & 95\% CI\\
\midrule
Sex & 0.71 & (0.41, 1.24) & 0.68 & (0.39, 1.19)\\
Disease & 0.55 & (0.31, 1) & 0.52 & (0.29, 0.93)\\
Phase (CR2 vs. CR1) & 1.18 & (0.47, 2.93) & 1.21 & (0.48, 3.02)\\
Phase (CR3 vs. CR1) & 1.47 & (0.38, 5.63) & 1.67 & (0.43, 6.5)\\
Phase (Relapse vs. CR1) & 4.30 & (2, 9.27) & 4.55 & (2.09, 9.9)\\
\addlinespace
Source & 1.56 & (0.52, 4.7) & 1.46 & (0.47, 4.54)\\
Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{CodeInput}
R> # Pick 100 equidistant points between 0 and 60 months
R> time_points <- seq(0, 60, length.out = 50)
R> 
R> # Data.frame containing risk profile
R> newdata <- data.frame(
R+   "Sex" = factor(c("F", "F"),
R+     levels = levels(bmtcrr[, "Sex"])
R+   ),
R+   "D" = c("ALL", "AML"), # Both diseases
R+   "Phase" = factor(c("Relapse", "Relapse"),
R+     levels = levels(bmtcrr[, "Phase"])
R+   ),
R+   "Age" = c(35, 35),
R+   "Source" = factor(c("PB", "PB"),
R+     levels = levels(bmtcrr[, "Source"])
R+   )
R+ )
R> 
R> # Estimate absolute risk curve
R> risk_cb <- absoluteRisk(
R+   object = model_cb, time = time_points,
R+   method = "numerical", newdata = newdata
R+ )
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(timereg)
R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) +
R+   const(Phase) + const(Source) + const(Age),
R+ data = bmtcrr, cause = 1, model = "fg"
R+ )
R> 
R> # Estimate absolute risk curve
R> risk_fg <- predict(model_fg, newdata, times = time_points)
\end{CodeInput}
\end{CodeChunk}

{\centering \includegraphics{../figures/bmtcrr-risk-1}

\caption{\label{fig:compAbsrisk} Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
\end{figure}
\end{CodeChunk}

@@

In the first two case studies, we described the basic functionalities of
the \pkg{casebase} package: creating population-time plots, fitting
parametric models for hazard functions, and estimating the corresponding
cumulative incidence curves. In the next two case studies, we explore
advanced features.

For the third case study, we will demonstrate regularization during
hazard estimation. Since we are using logistic regression, existing
packages, like \pkg{glmnet} can be leveraged. To demonstrate this
feature, we examine the Study to Understand Prognoses Preferences
Outcomes and Risks of Treatment (SUPPORT) dataset
\citep{knaus1995support}. The SUPPORT dataset tracks death for
individuals who are considered seriously ill within 5 hospitals in the
United States of America. Imputation was conducted using the \pkg{mice}
package in \proglang{R} with default settings \citep{mice}. The imputed
dataset is available as part of the \pkg{casebase} package. For more
information please see the SUPPORT data documentation in the package.

we have a total of 9104 observations and 30 variables, two of which are
SUPPORT day 3 physiology score (sps) and APACHE III day 3 physiology
score (aps). These scores were developed for a period of 180 days, which
makes our comparison over 5.56 years exploratory. Therefore, we will
assess these two models before using either as our baseline, by
comparing them to a full model using all variables. Also, our single
imputation introduces bias in the data. As the goal here is purely
demonstrative, this bias will be consistent with any of the following
models. As such, their comparison is still valid, though any
interpretation requires an awareness of said bias. A description of used
variables can be found in Table \ref{tab:support1} and a breakdown of
each categorical variable in Table \ref{tab:support2}. The original data
was obtained from the Department of Biostatistics at Vanderbilt
University \citep{harrell_2020}.

@@

Before generating an absolute risk curve using regularized logistic
regression, we will explore three options as a baseline comparison,
beyond a Kaplan-meier curve. The original study compared these scores
over a span of 180 days, which may not be comparable over the full 5.56
years. Therefore, We will examine three cox regression scenarios: sps
only, aps only and all covariates excluding sps and aps. Next, we will
compare each modeling using the Brier score \citep{graf1999ass}. The
Brier score is able to demonstrate both discrimination and calibration
in a single model \citep{graf1999ass} and provide a comparable metric
between semi-parametric and parametric methods in the survival context.
For all the following analyses in this case study, we will use 80\% of
the observations to train our models, while 20\% will be used to
generate absolute risk curves, all of which will be averaged to
demonstrate each model's predictive ability. Over 5.56 years,
\ref{fig:baselines} demonstrates that all the cox models have similar,
if not identical average predictions. Figure \ref{fig:brier1} examines
the brier scores as they change over time. As the full model has the
lowest brier score on overall, it will be used in the following absolute
risk comparison to regularized cox, regularized logistic regression and
the unadjusted Kaplan-Meier curve.

@@

{\centering \includegraphics{../figures/coefplots-1}
